---
title: "Educational Attainment in Los Angeles County"
subtitle: "A tract-level map of bachelor's degree attainment using ACS data"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidycensus)
library(leaflet)
library(tidyverse)
library(sf)
library(ggplot2)

census_key <- Sys.getenv("CENSUS_API_KEY")

if (identical(census_key, "")) {
  stop("Set the CENSUS_API_KEY environment variable before rendering this project.")
}

edu_data <- get_acs(
  geography = "tract",
  variables = c(
    bachelors = "B15003_022",
    total = "B15003_001"
  ),
  state = "CA",
  county = "Los Angeles",
  year = 2022,
  geometry = TRUE,
  survey = "acs5"
)

edu_wide <- edu_data %>%
  select(GEOID, NAME, variable, estimate, geometry) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(percent_bachelors = 100 * bachelors / total)
```

## Summary

This project maps bachelor's degree attainment across Los Angeles County census tracts using American Community Survey data.

The goal is to make educational inequality visible at the neighborhood level. County-wide averages can hide large local differences, but a tract-level map helps show where human-capital attainment is concentrated and where it may lag behind.

## Static Map

```{r}
#| fig-cap: "Bachelor's Degree Attainment in Los Angeles County Census Tracts"

ggplot(edu_wide) +
  geom_sf(aes(fill = percent_bachelors), color = NA) +
  scale_fill_viridis_c(option = "plasma", name = "% Bachelor's") +
  labs(
    title = "Educational Attainment Varies Sharply Across Los Angeles County",
    subtitle = "Share of residents with a bachelor's degree by census tract",
    caption = "Source: U.S. Census Bureau ACS 5-year estimates via tidycensus"
  ) +
  theme_minimal()
```

## Interactive Map

```{r}
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = edu_wide$percent_bachelors,
  na.color = "transparent"
)

leaflet(data = edu_wide) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~pal(percent_bachelors),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = ~paste0(
      NAME,
      "<br>% Bachelor's: ",
      round(percent_bachelors, 1),
      "%"
    ),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "13px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~percent_bachelors,
    opacity = 0.7,
    title = "% Bachelor's Degree",
    position = "bottomright"
  )
```

## Interpretation

Educational attainment is not evenly distributed across local space. Some tracts show very high bachelor's attainment, while others remain far below the county's better-performing areas.

That matters economically because educational attainment is closely tied to labor-market opportunity, household income, and long-run regional inequality. A map like this can serve as a starting point for deeper work on local opportunity gaps, school quality, housing, and neighborhood stratification.
